---
title: Seattle Bike Share Usage
author: Arman Bilge, Yanbo Ge, and Xiaoliu Wu
date: February 27
---

# Intro

<style type="text/css">
  .reveal h1 { font-size: 1.5em; }
  .reveal h2 { font-size: 1.25em; }
  .reveal { font-size: 36px; }
  img {
    width: 480px;
    height: 480px;
  }
</style>

```{r setup, echo = F, message = F}
knitr::opts_chunk$set(cache = F)
set.seed(666)
library(chron)
library(gplots)
library(httr)
library(jsonlite)
library(plyr)
API_KEY <- "AIzaSyAnu1sbEOUwfv2s4IfLyLYvSrn4ODMZSy4"
stations <- read.csv("2016_station_data.csv", fill = T, header = T)
trips <- read.csv("2016_trip_data.csv", fill = T, header = T)
weather <- read.csv("2016_weather_data.csv", fill = T, header = T)
census <- read.csv("neighborhood_variables.csv", fill = T, header = T)
```

```{r elevation, echo = F}
with.elevation <- function(df) {
  base.url <- "https://maps.googleapis.com/maps/api/elevation/json"
  positions <- apply(df, 1, function(row) paste(row["lat"], ",", row["long"], sep = ""))
  url <- httr:::handle_url(url = base.url, query = list(key = API_KEY, locations = paste(positions, collapse = "|")))$url
  df$elevation <- fromJSON(url)$results$elevation
  return(df)
}
stations <- with.elevation(stations)
```

```{r census, echo = F}
census$name <- NULL
stations <- merge(stations, census, by = "station_id")
```

```{r filters, echo = F}
members.only <- function(df) subset(df, usertype == "Member")
strip.times <- function(df) {
  df$starttime <- as.Date(df$starttime)
  df$stoptime <- as.Date(df$stoptime)
  return(df)
}
weekdays.only <- function(df) subset(df, weekend == F)
```

```{r datetime, echo = F}
parse.datetime <- function(df) {
  df$starttime <- strptime(as.character(df$starttime), "%m/%d/%Y %H:%M")
  df$stoptime <- strptime(as.character(df$stoptime), "%m/%d/%Y %H:%M")
  df$weekend <- is.weekend(df$starttime)
  return(df)
}
trips <- parse.datetime(trips)

parse.date <- function(df) {
  df$Date <- as.Date(as.character(df$Date), "%m/%d/%Y")
  return(df)
}
weather <- parse.date(weather)
```

```{r full_data, echo = F}
count.daily <- function(df) aggregate(list(trips = df$trip_id), by = list(date = df$starttime), FUN = length)
count.daily.station.inflow <- function(df) {
  df <- aggregate(list(trips = df$trip_id), by = list(date = df$starttime, to_station_id = df$to_station_id), FUN = length)
  product <- expand.grid(to_station_id = unique(df$to_station_id), date = unique(df$date))
  df <- merge(df, product, all = T)
  df$trips[is.na(df$trips)] <- 0
  return(df)
}
trips <- weekdays.only(strip.times(trips))
full.data <- merge(count.daily(trips), weather, by.x = "date", by.y = "Date")
inflow.data <- merge(merge(count.daily.station.inflow(trips), stations, by.x = "to_station_id", by.y = "station_id"), weather, by.x = "date", by.y = "Date")
```

<div style="width: 135%; height: 500px; overflow: hidden; position: relative"><iframe id="visual" src="http://willleahy.info/bikeshare/" style="width: 200%; height: 500px; position: absolute; left: -329px"></iframe></div>

# Question

Which features of a station affect its daily usage?

Can we use these to decide where to place stations?

# Data

## Sources

* Pronto! (October 2014–August 2016)
    * record for every trip taken
    * daily weather report
* Google Maps API
    * elevation
    * nearby points-of-interest
* Puget Sound Regional Council Census
    * socio-economic data

## Covariates of Interest

* elevation
* job density
* average income
* also hope to use:
    * number of nearby restaurants
    * number of bus routes at nearest stop

## Confounding Variables

* weather
    * temperature
    * precipitation
* total membership (not in data!)

# Exploration

## Temperature

```{r temperature, echo = F, fig.cap = ''}
plot(full.data$Mean_Temperature_F, full.data$trips)
abline(lm(trips ~ Mean_Temperature_F, data = full.data), col = "red")
```

## Precipitation

```{r precipitation, echo = F, fig.cap = ''}
plot(full.data$Precipitation_In, full.data$trips)
abline(lm(trips ~ Precipitation_In, data = full.data), col = "red")
```

## Estimates

```{r estimates, echo = F}
summary(lm(trips ~ Mean_Temperature_F + Precipitation_In, data = full.data))
```

# Methods

##

* Dropped weekends
* Counted number of daily arrivals per station
* Fit linear model with weather and covariates of interest
* Two independent code bases give same results

## Model

$$
y \sim \beta X + \epsilon, \epsilon \sim \mathcal{N}(0, \sigma I)
$$

* $y$ is daily "in-flow" per station
* $X$ is design matrix including covariates and confounding variables
* no interaction terms (for now)

# Initial Analysis

```{r initial.analysis, echo = F}
summary(lm(trips ~ elevation + jobden + income + Mean_Temperature_F + Precipitation_In, data = inflow.data))
```

# Challenges

## Dependencies between stations

* Total number of bikes remains constant
* Complex network dependencies between stations (not good for linear regression)
* Some stations were installed later on
* Solution: find a subset of stations that seem independent
    * Create a correlation matrix of residuals after regressing on confounding variables

## Heatmap

```{r heatmap, echo = F, fig.cap = ''}
inflow.complete <- inflow.data[!is.na(inflow.data$Mean_Temperature_F) & !is.na(inflow.data$Precipitation_In),]
model <- lm(trips ~ Mean_Temperature_F + Precipitation_In, data = inflow.complete)
inflow.complete$residual <- residuals(model)
x <- unstack(inflow.complete[,c("residual", "to_station_id")])
x$"DPD-03" <- NULL
cor.matrix <- cor(data.frame(x))
heatmap.2(cor.matrix, trace = "none", dendrogram = "none")
```

## Total membership

* Data not included in public dataset
* Estimate a fixed effect representing total membership for each month?

# Concluding Remarks

* Temperature and precipitation significantly explain variation
* Coefficients for elevation and socio-economic factors are significant and intuitive
* Possible method for identifying independent stations

# Questions?

Thanks for listening!
